---
author: |
  | William Findlay
title: |
  | Honours Thesis Pre-Proposal
subtitle: |
  | The Viability of eBPF for
  | Intrusion Detection Implementations
date: \today
bibliography: /home/housedhorse/.bibs/uni.bib
csl: /home/housedhorse/.bibs/ieee.csl
subparagraph: yes
header-includes:
    - \usepackage{findlayrmd}
output:
    pdf_document:
      number_sections: true
      fig_crop: true
      fig_caption: true
      keep_tex: false
    word_document: default
---
```{r,include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
library(lubridate)
```
\pagestyle{fancy}
\thispagestyle{plain}
\counterwithin{lstlisting}{section}
\includepdf{cover-page-filled.pdf}
\renewcommand{\maketitle}{\oldmaketitle}
\setcounter{page}{1}
\maketitle

# Overview of Thesis

This section presents an overview of the thesis including motivation, objectives, research questions,
and methodology.

## Motivation

Modern computer systems are inherently complex; this complexity yields a plethora of opportunities for malicious actors
to exploit such systems to their benefit. *Intrusion detection systems* seek to solve the problem of unwanted behavior
through a variety of techniques, ranging from relatively simple heuristics, to neural networks.

Recent patches to the Linux Kernel have brought about *Extended Berkeley Packet Filtering* (*eBPF*), a technology
which offers a myriad of prospects for use in intrusion detection systems. eBPF programs run in an isolated virtual machine,
which means that they are able to hook into kernel-level tracing APIs while offering unprecedented stability -- an eBPF program
is guaranteed to never crash the kernel, and a built-in verifier prevents the loading of eBPF programs which could potentially damage
the system or produce unwanted side effects.
Using eBPF, it is possible to build a robust, portable, and future-proof intrusion detection system that leverages existing
techniques and empowers them with a highly performant, modern, and safe kernel tracing API.

*Process Homeostasis* (*pH*), introduced in a 2002 dissertation by Dr. Anil Somayaji, is a perfect example of
an intrusion detection system that can greatly benefit from the boons offered by eBPF. While pH offered a highly effective
and compelling solution at the time, it suffered from a lack of official upstream support and, as it was implemented as a patch for Linux 2.2,
soon became outdated. The advent of eBPF presents a unique opportunity to bring pH back and allow it to compete with
modern solutions -- many of which entail a much higher overhead, as well as a higher propensity to yield false positives in practice.

The proposed system, styled as *ebpH* (*Extended Berkeley Process Homeostasis*), builds on its predecessor, maintaining core functionality
while offering the advantages of eBPF, including improvements to stability, flexibility, and widespread adoptability thanks to
the prevalence of eBPF in the modern Linux world.

## Main Objectives

While eBPF offers great promise to the area of intrusion detection and indeed computer security as a whole,
much yet remains unexplored with respect to its feasibility in such applications. As such, the primary goal of
this project will be to ascertain the degree of viability (or lack thereof) of eBPF implementations for production-grade
IDS software.

In order to test the its viability for such applications, eBPF will be used to implement *ebpH*, a program that attempts to
encompass the primary functionality of the original pH software by Dr. Somayaji. This means that it should be able
to construct and maintain system-wide profiles for every binary, as well as detect and notify the end user of anomalous system
calls as they occur throughout the system. Additional work will be done to make small improvements and/or changes to
the original design where necessary.

For further details regarding the testing of ebpH (including criteria for performance, efficacy, and overhead), please refer to \autoref{methodology} of this document.

## Research Questions

This thesis is primarily concerned with the viability of eBPF-based intrusion detection systems.
In particular, is such an implementation possible, how might it be done, and why is it worth doing?
As such, the primary research questions which will be explored are as follows:

- What factors might contribute to eBPF being ideal for the implementation of an intrusion detection system?
  - Do other methods of implementation already provide these benefits?
- What caveats (if any) might exist which detract from eBPF's use for such implementations?
  - Is it possible to work around said caveats if necessary?
- How might an eBPF-based intrusion detection system perform (see \autoref{methodology})?

## Methodology

In order to ascertain whether eBPF is a viable method for implementing an intrusion detection system,
ebpH will be implemented and tested for the following:

- Coverage: What types of attacks can the system detect and/or prevent?
- Overhead: What performance penalties are observed while running under various loads?
- Limitations: Is there anything another implementation can do that an eBPF implementation cannot?
  - Conversely, can an eBPF implementation circumvent any limitations imposed by other implementations?

ebpH will be tested in a controlled, virtualized environment, using conventional software testing methods.
Additional testing will be done in a production or quasi-production environment.

# Bi-Weekly Schedule for First Term (Fall 2019)

\autoref{fall-sched} presents a bi-weekly schedule for the fall 2019 term. This term will be used
primarily for background research, implementation of various eBPF programs, and developing a plan for testing in the winter term.

<!-- print table for bi-weekly fall schedule -->

```{r}
# weekly
dates = c(seq(as.Date("2019-09-04"), as.Date("2019-12-06"), by = 'week'))

# biweekly
start_dates = dates[seq(1,length(dates),2)]
end_dates = start_dates + 13
end_dates[7] = as.Date("2019-12-06")

dates = paste(format(start_dates, "%b %d"), "-", format(end_dates, "%b %d, %Y"))
tasks = NULL

for(i in 1 : length(dates)) {
  tasks = c(tasks,NA)
}

tasks[1] = "Background research on various intrusion detection methods."
tasks[2] = "Background research on bcc (BPF compiler collection) and eBPF."
tasks[3] = "Implementation of ebpH."
tasks[4] = "Implementation of ebpH."
tasks[5] = "Implementation of ebpH."
tasks[6] = "Draft thesis proposal report."
tasks[7] = "Finalize thesis proposal report."

dates[7] = paste(dates[7],footnote_marker_alphabet(1))

df = data.frame(week=dates,task=tasks)

names(df) = c("Bi-weekly Period","Task")

df %>%
  kable(longtable=F, align = c('r','l'), caption=('\\label{fall-sched}A bi-weekly schedule for the fall 2019 term.'), escape=F) %>%
  kable_styling(latex_options=c('striped','hold_position')) %>%
  row_spec(0,bold=T) %>%
  column_spec(2,width='30em') %>%
  add_footnote("This period is slightly shorter to reflect the end of the term.")
```
